\bf On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2024-11-24 Manuscript/manuscript.tex
smoothed results section

Let $Y_{ijk}$ denote the reaction time of observation~$k$ in heat~$j$
of year~$i$. Conditioning on a venue effect $v_i$ for year~$i$ 
and a heat effect $h_{i/j}$ nested within each year~$i$, the
distribution of $Y_{ijk}$ is
$\text{GG}(\mu_{ijk}, \sigma_{ijk}, \nu)$, where
\begin{align}
\log(\mu_{ijk}) &= \beta_0 + v_i ,\\
\log(\sigma_{ijk}) &= \gamma_0 + h_{i/j},
\end{align}
$v_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $h_{i/j}$ is normally distributed with mean
zero and variance~$\sigma_h^2$.
The two random effects were found useful: one capturing the venue effect, which
is used to contrast years, and the second being the heat effect, where every
race was given a unique identifier with typically five to nine observations
per race. The heat effect is important as it captures the variability in the
amount of time athletes are on the starting blocks before the gun goes off.
This model can be fit with R package \texttt{gamlss}
\citep{stasinopoulos2024generalized}.

@@

\begin{table}
  \centering
  \caption{Summaries of the results of comparisons between 
    reaction times from different competitions for the same athletes. 
    2022 Nat. vs Inter. compares reaction times from 2022 national-level 
    championships and the 2022 World Track and Field Championships. 2019 
    vs 2022 compares reaction times from the 2019 and 2022 World Track and 
    Field Championships. 2022 vs 2023 compares reaction times from the 
    2022 and 2023 World Track and Field Championships.}
  \begin{tabular}{c c c c c} 
   \toprule
   Comparison & Permutation & Asymptotic & Athletes & Number observations  \\ 
   \midrule
   2022 Nat. vs Inter. Men & $1.0 \cdot 10^{-6}$ & $ 6.1 \cdot 10^{-5}$ & 17 & 80 \\
   2022 Nat. vs Inter. Women & $1.0 \cdot 10^{-6}$ & $ 1.2 \cdot 10^{-3}$ & 17 & 80 \\[1ex]
   2019 vs 2022 Men & $2.8 \cdot 10^{-5}$ & $1.1 \cdot 10^{-5}$ & 34 & 134 \\
   2019 vs 2022 Female & $ 1.5 \cdot 10^{-3}$ & $6.9 \cdot 10^{-3}$ & 31 & 124 \\[1ex]
   2022 vs 2023 Men & $1.0 \cdot 10^{-6}$ & $1.4 \cdot 10^{-6}$ & 45 & 161 \\
   2022 vs 2023 Female & $1.0 \cdot 10^{-6}$ & $9.4 \cdot 10^{-7}$ & 47 & 182 \\
   \bottomrule
  \end{tabular}
  \label{tab:Clusrankresults}
\end{table}

@@

We considered many types of models from various continuous distribution families
before ultimately deciding on the generalized Gamma model desrcribed in
Section~\ref{sec:gamlss}.  The residuals of the GG model were particularly
strong, as evidenced by Figure~\ref{fig:Residuals}.  We care especially about
the lower residuals as we are trying to estimate the probability of a low reaction
time.  While there are two outliers present in the residual plot, we accept that
this is relatively small in relation to the rest of the roughly 770 observations.

\begin{figure}[tbp]
  \centering
  \includegraphics{diagnosis.pdf}
  \caption{Diagnosis of the fitted generalized Gamma distribution with
    random effects in model parameters: kernerl density of 1 million
    observations drawn from the fitted model overlaid with the
    histogram of the observed reaction times (left); Q-Q plot of the
    z-score of the quantile residuals from the fitted model.}
  \label{fig:diagnosis}
\end{figure}

\begin{table}
  \centering
  \caption{AIC and estimated parameters for the gg3b model fit}
  \label{tab:Gamma_parameters}
  \begin{tabular}{c c c c c c c}
    \toprule
    Data set & AIC & $\mu$ & $\sigma$ & $\nu$ & $\sigma_v$ & $\sigma_h$ \\
    \midrule
    Excluding 2022 & -3660 & -1.900 & -2.192 & -1.315 & 0.0434 & 0.3259 \\
    Including 2022 & -4103 & -1.910 & -2.200 & -1.178 & 0.0576 & 0.3198 \\
    \bottomrule
  \end{tabular}
\end{table}

\of{Created these with the function I wrote model-name at bottom of gamlss.R. Do
  any other parametes need to be added?}
\jy{Remove AIC because there is no comparison. Need their standard errors.}

The estimated parameters of the GLMM model under the log link are shown in 
Table~\ref{tab:Gamma_parameters}, both excluding and including 2022 data.
.  We report the AIC, the point estimates for $\mu$,
$\sigma$, and $\mu$, as well as the standard deviations of the venue and heat
effects: $\sigma_v$ and $\sigma_h$ respectively. When we exclude 2022 from the
full model, we observe small but intriguing discrepancies in the models. The
standard deviation of the venue effect, $\sigma_v$, increases from 0.0434 to
0.0576 when 2022 is included. This suggests that by including 2022 in the model,
there is more of a variation in venue effects.

\begin{figure}[tbp]
  \centering
  \includegraphics{ComparisonOfVenueEffects}
  \caption{The venue effects from 1999 to 2023 estimated from the GLMM for the
    data including 2022 (top) and excluding 2022 (bottom).}
  \label{fig:VenueEffects}
\end{figure}

The uniqueness of 2022 is further demonstrated by Figure~\ref{fig:VenueEffects},
which shows the values of the venue random effects.  The temporal patterns
of the two sets of effects look very similar, except that the positive effects
estimated with 2022 included have higher magnitude than those estimated without
2022, which compensates the negative effect of the 2022 with a large magnitude.
Both plots agree to some extent with the pattern of the boxplots in
Figure~\ref{fig:Boxplot}. The mismatch is due to the heat effect, since the venue
effect only captures a smaller part of the variation than the heat effect as
evident from their standard deviations. It is possible to calculate the extremity
of the 2022 heat effect as we know that the mean of the heat effects is $0$,
the standard deviation is $0.42$ as given by Table~\ref{tab:Gamma_parameters},
and the estimate of the 2022 heat effect is $-0.93$. We find the tail probability
to be $0.0128$ indicating there is a low chance of observing a venue effect as
extreme as 2022.

\begin{figure}[tbp]
  \centering
  \includegraphics{GGDensity}
  \caption{The densities of the GG distribution with random effects for heat
  and venue (black), and without random effects (blue)}
  \label{fig:DensityPlot}
\end{figure}
\of{Not sure what to write about the density plot or how to fit it into the rest
of the section.}

\begin{table}
  \centering
  \caption{Probabilities of observing reaction times less than threshold 0.08,
  0.09, and 0.10 seconds based on the two effect model.}
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & Threshold 0.08 & Threshold 0.09 & Threshold 0.10  \\ 
   \midrule
   Excluding 2022 & $5.31\cdot10^{-5}$ & $3.53\cdot10^{-4}$ &  $1.94\cdot10^{-3}$  \\ 
   Including 2022 & $6.84\cdot10^{-5}$ & $4.95\cdot10^{-4}$ & $2.76\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_probability}
\end{table}
\of{table has been updated with gg3b results}
\jy{Could you also test sensitivity for excluding the positive but
  fault reaction times?}

The GLMM with both venue and heat random effects helps us to assess how extreme
a reaction time below a given threshold is. The probability of observing a reaction
time below a threshold, assuming no intentional false starts, can be
approximated by generating a large number of realizations from the fitted
GLMM. We used 10 millions realizations in the analysis. 
Table~\ref{tab:Sim_probability} summarizes the estimated probabilities of
observing a reaction time below 0.08, 0.09, and 0.10 seconds under two different
scenarios: one excluding and the other including data from 2022. Apparently,
excluding 2022 lowered the probability of observing a fast reaction time, but
the difference is not big in magnitude. For example, the probability for
seeing a reaction time below 0.10 seconds changes from $2.36\cdot 10^{-3}$ to
$2.90\cdot 10^{-3}$ when 2022 is included. These values correspond to one in
every 344 starts and 423 starts, respectively. The table shows that changing the
reaction time barrier from 0.1 seconds to 0.08 seconds drastically reduces the
chance of observing a reaction time below the barrier, from one in roughly every
283 starts to one in every 58480 starts using data with 2022 included. The
results substantiates the recommendations by \citet{komi2009iaaf}.

\begin{table}
  \centering
  \caption{Suggested reaction time barriers based on tail probabilities.}
  \begin{tabular}{c c c} 
   \toprule
   Data Set & Tail probability  $10^{-3}$ & Tail probability $10^{-4}$ \\ 
   \midrule
   Excluding 2022 & $0.096$ & $0.083$ \\ 
   Including 2022 & $0.094$ & $0.082$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_time}
\end{table}
\of{table has been updated with gg3b results}

Utilizing the same model, we can subtly shift our interpretation to determine a
suitable reaction time barrier based on the probability of observing a time
below said barrier. Including 2022 data, a reaction time barrier of 0.092
seconds is warranted to maintain a 0.1\% chance of observing an illicitly fast
reaction time, as delineated in Table~\ref{tab:Sim_time}. This calculation can
be replicated for various probability levels; for instance, a probability of
0.0001 (equating to one disqualification per 10,000 starts) necessitates a
reaction time barrier of 0.083 seconds. The exclusion of 2022 data, which
inherently lowers the probability of observing a swift reaction time, mandates a
less lenient reaction time barrier when calculated based on probability, albeit
without a substantial difference in magnitude.

@@

This paper aspires to offer a statistical lens through which to examine Allen's
disqualification at the 2022 World Track and Field Championships, rather than
delivering a conclusive judgment regarding potential equipment malfunction. Our
findings reveal that athletes' reaction times at the event were, in general,
faster than those recorded at other competitions, with
Table~\ref{tab:Clusrankresults} presenting multiple significant p-values that
highlight disparities in average reaction times between the 2022 Championship
and other competitions. Further, our GLMM results indicate that a reaction time
of 0.1 seconds may not be as exceptional as commonly believed. According to the
probabilities in Table~\ref{tab:Sim_time}, World Athletics might contemplate
adjusting the disqualification barrier to 0.08 seconds, thereby enabling athletes
like Allen to react more swiftly without the apprehension of
disqualification. In summary, while the results designate 2022 as an anomalous
year, Allen's time, despite resulting in disqualification, may not be
categorically extreme.

>>>>

Let $Y_{ijk}$ denote the reaction time of observation~$k$ in heat~$j$
of year~$i$. Conditioning on a venue effect $v_i$ for year~$i$ 
and a heat effect $h_{i/j}$ nested within each year~$i$, the
distribution of $Y_{ijk}$ is
$\text{GG}(\mu_{ijk}, \sigma_{ijk}, \nu)$, where
\begin{align}
\log(\mu_{ijk}) &= \beta_0 + v_i , \label{eq:mu}\\
\log(\sigma_{ijk}) &= \gamma_0 + h_{i/j} , \label{eq:sigma}
\end{align}
$v_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $h_{i/j}$ is normally distributed with mean
zero and variance~$\sigma_h^2$.
The two random effects were found useful: one capturing the venue effect, which
is used to contrast years, and the second being the heat effect, where every
race was given a unique identifier with typically five to nine observations
per race. The heat effect is important as it captures the variability in the
amount of time athletes are on the starting blocks before the gun goes off.
This model can be fit with R package \texttt{gamlss}
\citep{stasinopoulos2024generalized}.

@@

\begin{table}
  \centering
  \caption{P-values of comparisons between 
    reaction times from different competitions for the same athletes. 
    2022 Nat. vs Inter. compares reaction times from 2022 national-level 
    championships and the 2022 World Track and Field Championships. 2019 
    vs 2022 compares reaction times from the 2019 and 2022 World Track and 
    Field Championships. 2022 vs 2023 compares reaction times from the 
    2022 and 2023 World Track and Field Championships.}
  \begin{tabular}{c c c c c} 
   \toprule
   Comparison & Permutation & Asymptotic & \# of athletes & \# of observations  \\ 
   \midrule
   2022 Nat. vs Inter. Men & $1.0 \cdot 10^{-6}$ & $ 6.1 \cdot 10^{-5}$ & 17 & 80 \\
   2022 Nat. vs Inter. Women & $1.0 \cdot 10^{-6}$ & $ 1.2 \cdot 10^{-3}$ & 17 & 80 \\[1ex]
   2019 vs 2022 Men & $2.8 \cdot 10^{-5}$ & $1.1 \cdot 10^{-5}$ & 34 & 134 \\
   2019 vs 2022 Female & $ 1.5 \cdot 10^{-3}$ & $6.9 \cdot 10^{-3}$ & 31 & 124 \\[1ex]
   2022 vs 2023 Men & $1.0 \cdot 10^{-6}$ & $1.4 \cdot 10^{-6}$ & 45 & 161 \\
   2022 vs 2023 Female & $1.0 \cdot 10^{-6}$ & $9.4 \cdot 10^{-7}$ & 47 & 182 \\
   \bottomrule
  \end{tabular}
  \label{tab:Clusrankresults}
\end{table}

@@

\jy{Please add standard errors, which could be from \texttt{summary(gg3b)}}

\jy{Are these numbers correct? I modified with the gg3b in gg.R}

\begin{table}
  \centering
  \caption{Estimated fixed-effect parameters with standard errors in
    parentheses and estimated variance of the random effects from the
    fitted generalized Gamma distribution with venue level random
    effects in $\mu$ and heat level random effects in $\sigma$ in
    Models~\eqref{eq:gg}--\eqref{eq:sigma}.}
  \label{tab:ggfit}
  \begin{tabular}{c c c c c c}
    \toprule
    Data set & $\beta_0$ & $\gamma_0$ & $\nu$ & $\sigma_v$ & $\sigma_h$ \\
    \midrule
    Excluding 2022 & $-$1.910 (0.005)  & $-$2.200 (0.025) & $-$1.177 (0.442) & 0.043 & 0.326 \\
    Including 2022 & $-$1.910 & $-$2.200 & $-$1.178 & 0.058 & 0.320 \\
    \bottomrule
  \end{tabular}
\end{table}

The fitted parameters of the generalized Gamma distribution int he
GAMLSS framework in Equations~\eqref{eq:gg}--\eqref{eq:sigma} are
summarized in Table~\ref{tab:ggfit}. Results obtained from both
excluding and including 2022 data are reported. The fixed-effect
parameters include $\beta_0$, $\gamma_0$, and $\nu$, corresponding to
the intercept of the log-location, log-scale, and shape of the GG
distribution, respectively. Random effects account for variability at
the venue level $\sigma_v$ on the log-scale of the $\mu$ parameter
and at the heat level $\sigma_h$ on the log-scale of the $\sigma$
parameter in the density in Equation~\eqref{eq:gg}. The variance 
of the venue-level random effect is smaller than the heat-level random
effect variance, suggesting that heat-level variability in the scale
parameter is substantial, though on the dispersion parameter. When the
2022 data is included, the fixed-effect estimates remain stable, 
indicating robustness, but the venue-level random-effect variance
increases to 0.0576, reflecting additional variability in the location
parameter, while the heat-level random-effect variance remains stable.
decreases slightly to 0.3198. These results
highlight that reaction times are influenced by both venue and heat-level 
factors, with heat-level variability playing a more prominent role in the scale 
parameter, and that the inclusion of 2022 introduces greater venue-level 
variability, likely due to systematic differences in reaction times that year.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{diagnosis.pdf}
  \caption{Diagnosis of the fitted generalized Gamma distribution with
    random effects in model parameters: kernerl density of 1 million
    observations drawn from the fitted model overlaid with the
    histogram of the observed reaction times (left); Q-Q plot of the
    normal z-score of the quantile residuals from the fitted model.}
  \label{fig:diagnosis}
\end{figure}

Figure~\ref{fig:diagnosis} presents diagnostic checks for the fitted 
generalized Gamma distribution model with random effects. The left 
panel compares the kernel density estimate of one million simulated reaction 
times from the fitted model to the histogram of the observed reaction times. 
The close alignment between the density curve and the histogram suggests that 
the fitted GG model adequately captures the overall distribution of the 
reaction times. The right panel shows a Q-Q plot of the z-scores of the 
quantile residuals from the fitted model. The points lie approximately along 
the 45-degree reference line, indicating that the residuals are consistent 
with the standard normal distribution, supporting the adequacy of the model 
fit. These diagnostics collectively demonstrate that the fitted model provides 
a reasonable representation of the observed reaction time data.

% \begin{figure}[tbp]
%   \centering
%   \includegraphics{ComparisonOfVenueEffects}
%   \caption{The venue effects from 1999 to 2023 estimated from the GLMM for the
%     data including 2022 (top) and excluding 2022 (bottom).}
%   \label{fig:VenueEffects}
% \end{figure}

% The uniqueness of 2022 is further demonstrated by Figure~\ref{fig:VenueEffects},
% which shows the values of the venue random effects.  The temporal patterns
% of the two sets of effects look very similar, except that the positive effects
% estimated with 2022 included have higher magnitude than those estimated without
% 2022, which compensates the negative effect of the 2022 with a large magnitude.
% Both plots agree to some extent with the pattern of the boxplots in
% Figure~\ref{fig:Boxplot}. The mismatch is due to the heat effect, since the venue
% effect only captures a smaller part of the variation than the heat effect as
% evident from their standard deviations. It is possible to calculate the extremity
% of the 2022 heat effect as we know that the mean of the heat effects is $0$,
% the standard deviation is $0.42$ as given by Table~\ref{tab:Gamma_parameters},
% and the estimate of the 2022 heat effect is $-0.93$. We find the tail probability
% to be $0.0128$ indicating there is a low chance of observing a venue effect as
% extreme as 2022.

\jy{Could you also test sensitivity for excluding the positive but 
  fault reaction times?}

\begin{table}
  \centering
  \caption{Probabilities of observing reaction times less than threshold 0.08,
  0.09, and 0.10 seconds based on the two effect model.}
  \begin{tabular}{c c c c} 
   \toprule
   Data Set & Threshold 0.08 & Threshold 0.09 & Threshold 0.10  \\ 
   \midrule
   Excluding 2022 & $5.31\cdot10^{-5}$ & $3.53\cdot10^{-4}$ &  $1.94\cdot10^{-3}$  \\ 
   Including 2022 & $6.84\cdot10^{-5}$ & $4.95\cdot10^{-4}$ & $2.76\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_probability}
\end{table}

The fitted generalized Gamma GAMLSS model with both venue- and heat-level 
random effects provides a framework for assessing how extreme reaction times 
below certain thresholds are. The probability of observing a reaction time 
below a given threshold, assuming no intentional false starts, was approximated 
by generating 10 million realizations from the fitted model. 
Table~\ref{tab:Sim_probability} summarizes the probabilities of observing 
reaction times below 0.08, 0.09, and 0.10 seconds under two scenarios: one 
excluding and the other including data from 2022. Excluding 2022 slightly 
reduces the probability of observing a fast reaction time, but the difference 
is small. For example, the probability of a reaction time below 0.10 seconds 
decreases from $2.76 \cdot 10^{-3}$ (approximately one in 362 starts) to 
$1.94 \cdot 10^{-3}$ (approximately one in 515 starts) when 2022 is excluded. 
Lowering the reaction time threshold from 0.10 to 0.08 seconds drastically 
reduces the likelihood of observing a reaction time below the barrier, with 
the probability dropping from one in every 362 starts (at 0.10 seconds) to one 
in every 14620 starts (at 0.09 seconds) and one in every 146198 starts (at 0.08 
seconds) when 2022 is included. These results highlight the rarity of extremely 
fast reaction times and substantiate the recommendations of \citet{komi2009iaaf} 
to carefully consider the selection of reaction time thresholds.

\begin{table}
  \centering
  \caption{Suggested reaction time barriers based on tail probabilities.}
  \begin{tabular}{c c c} 
   \toprule
   Data Set & Tail probability  $10^{-3}$ & Tail probability $10^{-4}$ \\ 
   \midrule
   Excluding 2022 & $0.096$ & $0.083$ \\ 
   Including 2022 & $0.094$ & $0.082$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_time}
\end{table}

Utilizing the same model, we can determine suitable reaction time barriers 
based on the probability of observing a time below the barrier. As shown in 
Table~\ref{tab:Sim_time}, including the 2022 data suggests a reaction time 
barrier of 0.094 seconds to maintain a 0.1\% chance of observing an 
exceptionally fast reaction time, while a stricter threshold of 0.082 seconds 
is needed to limit this probability to 0.01\%. Excluding the 2022 data results 
in slightly higher thresholds of 0.096 and 0.083 seconds for the respective 
probability levels. These results indicate that while the inclusion of 2022 
data slightly reduces the recommended barrier, the magnitude of the difference 
is relatively small. This approach allows for tailoring reaction time 
thresholds to desired levels of false positive rates, balancing fairness and 
precision in disqualification criteria.

@@

This paper aspires to offer a statistical lens through which to examine Allen's
disqualification at the 2022 World Track and Field Championships, rather than
delivering a conclusive judgment regarding potential equipment malfunction. Our
findings reveal that athletes' reaction times at the event were, in general,
faster than those recorded at other competitions, with
Table~\ref{tab:Clusrankresults} presenting multiple significant p-values that
highlight disparities in average reaction times between the 2022 Championship
and other competitions. Further, our GLMM results indicate that a reaction time
of 0.1-second may not be as exceptional as commonly believed. According to the
probabilities in Table~\ref{tab:Sim_time}, World Athletics might contemplate
adjusting the disqualification barrier to 0.08 seconds, thereby enabling athletes
like Allen to react more swiftly without the apprehension of
disqualification. In summary, while the results designate 2022 as an anomalous
year, Allen's time, despite resulting in disqualification, may not be
categorically extreme.

### 2024-11-24 Manuscript/reply.tex
smoothed results section

Thank you for the feedback. The purpose of Figure~3 is to display the venue 
effects for each year and visually illustrate how 2022 compares to other years. 
To address your concern, we added additional comments quantifying the 
probability of observing an extreme venue effect, which strengthens the 
argument that 2022 is anomalous. If needed, we are open to revising the plot 
further or integrating it into a single panel showing deltas, as suggested.
\eds{let's revisit this again after objective 2 analyses are set}

>>>>

Given the confusion it may cause, this figure has been removed.

% Thank you for the feedback. The purpose of Figure~3 is to display the venue 
% effects for each year and visually illustrate how 2022 compares to other years. 
% To address your concern, we added additional comments quantifying the 
% probability of observing an extreme venue effect, which strengthens the 
% argument that 2022 is anomalous. If needed, we are open to revising the plot 
% further or integrating it into a single panel showing deltas, as suggested.
% \eds{let's revisit this again after objective 2 analyses are set}
